Dissertation

### 2024-04-21 Manuscript/manuscript.qmd
small typo fixes

Varied training has been shown to influence learning in wide array of different tasks and domains, including categorization [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968; @nosofskyModelguidedSearchOptimal2019; @morgensternOneshotCategorizationNovel2019; @plebanekEffectsFrequencyVariability2021], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012; @brekelmansDoesHighVariability2022], anagram completion [@goodeSuperiorityVariableRepeated2008], perceptual learning [@manentiVariabilityTrainingUnlocks2023; @lovibondStimulusDiscriminabilityInduction2020; @robsonSpecificVariedPractice2022a; @zamanPerceptualVariabilityImplications2021], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], cognitive control tasks [@sabahWhenLessMore2019; @moshon-cohenStimulusVariabilityImproves2024], associative learning [@leeEvidentialDiversityIncreases2019; @reichmannVariabilityAbstractionEvaluative2023; @fanStimulusDiversityIncreases2022; @pradaExperiencedCategoryVariability2020; @liveseyRevisitingPeakShift2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], face recognition [@honigPerceptualSimilarityModulates2022; @burtonIdentityVariationRepresentations2016; @menonVariationPhotosSame2015],  the perception of social group heterogeneity [@linvilleExemplarAbstractionModels1993; @konovalovaInformationSamplingExplanation2020; @parkPerceptionVariabilityCategory1987; @gershmanStructureLearningPrinciples2023] ,  simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019; @breslinConstantVariablePractice2012], and  complex skill learning [@seowTransferEffectsVaried2019; @huetEducationAttentionExplanation2011; @hacquesVisualControlClimbing2022]. See @czyzVariabilityPracticeInformation2021 or @ravivHowVariabilityShapes2022 for more detailed reviews.

Research on the effects of varied training typically manipulates variability in one of two ways. In the first approach, a high variability group is exposed to a greater number of unique instances during training, while a low variability group receives fewer unique instances with more repetitions. Alternatively, both groups may receive the same number of unique instances, but the high variability group's instances are more widely distributed or spread out in the relevant psychological space, while the low variability group's instances are clustered more tightly together. Researchers then compare the training groups in terms of their performance during the training phase, as well as their generalization performance during a testing phase. Researchers will usually compare the performance of the two groups during both the training phase and a subsequent testing phase. The primary theoretical interest is often to assess the influence of training variability on generalization to novel testing items or conditions. However, the test may also include some or all of the items that were used during the training stage, allowing for an assessment of whether the variability manipulation influenced the learning of the trained items themselves, or to easily measure how much performance degrades as a function of how far away testing items are from the training items.

The influence of training variation has received a large amount of attention in the domain of sensorimotor skill learning. Much of this research has been influenced by the work of @schmidtSchemaTheoryDiscrete1975, who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. Schema theory presumes that learners possess general motor programs for a class of movements (e.g. an underhand throw). When called up for use motor programs are parameterized by schema rules which determine how the motor program is parameterized or scaled to the particular demands of the current task. Schema theory predicts that variable training facilitates the formation of more robust schemas, which will result in improved generalization or transfer. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices with multiple instances along some task-relevant dimension that remains invariant for the constant group. For example, studies using a projectile throwing task might assign participants to either constant training that practice throwing from a single location, or to a varied group that throws from multiple locations. Following training, both groups are then tested from novel throwing locations [@pigottMotorSchemaStructure1984; @willeyLimitedGeneralizationVaried2018; @pachecoLearningSpecificIndividual2018; @wulfEffectTypePractice1991].

One of the earliest, and still often cited investigations of Schmidt's benefits of variability hypothesis was the work of @kerrSpecificVariedPractice1978. Two groups of children, aged 8 and 12, were assigned to either constant or varied training of a bean bag throwing task. The constant group practiced throwing a bean-bag at a small target placed 3 feet in front of them, and the varied group practiced throwing from a distance of both 2 feet and 4 feet. Participants were blindfolded and unable to see the target while making each throw but would receive feedback by looking at where the beanbag had landed in between each training trial. 12 weeks later, all of the children were given a final test from a distance of 3 feet which was novel for the varied participants and repeated for the constant participants. Participants were also blindfolded for testing and did not receive trial by trial feedback in this stage. In both age groups, participants performed significantly better in the varied condition than the constant condition, though the effect was larger for the younger, 8-year-old children. This result offers a particularly compelling example of the merits of varied practice, given that the varied group was able to outperform the constant group even from the  home turf location where one may have expected the constatn group to have the strongest advantage. A similar pattern of results was observed in another study wherein varied participants trained with tennis, squash, badminton, and short-tennis rackets were compared against constant subjects trained with only a tennis racket [@greenPracticeVariabilityTransfer1995a]. One of the testing conditions had subjects repeat the use of the tennis racket, which had been used on all 128 training trials for the constant group, and only 32 training trials for the varied group. Nevertheless, the varied group outperformed the constant group when using the tennis racket at testing, and also performed better in conditions with several novel racket lengths. However, as is the case with many of the patterns commonly observed in the "benefits of variability" literature, the pattern wherein the varied group outperfroms the constant group even from the constants group's home turf has not been consistently replicated. One recent study attempted a near replication of the  Kerr & Booth study [@willeyLongtermMotorLearning2018], having subjects throw beanbags at a target, with the varied group training from positions (5 and 9 feet) on either side of the constant group (7 feet). This study did not find a varied advantage from the constant training position, though the varied group did perform better at distances novel to both groups. However, this study diverged from the original in that the participants were adults; and the amount of training was much greater (20 sessions with 60 practice trials each, spread out over 5-7 weeks).

@@

Benefits of varied training have also been observed in many studies outside of the sensorimotor domain.  @goodeSuperiorityVariableRepeated2008 trained participants to solve anagrams of 40 different words ranging in length from 5 to 11 letters, with an anagram of each word repeated 3 times throughout training, for a total of 120 training trials. Although subjects in all conditions were exposed to the same 40 unique words (i.e. the solution to an anagram), participants in the varied group saw 3 different arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for the solution word FLOOD, whereas constant subjects would train on three repetitions of LDOOF (spread evenly across training). Two different constant groups were used. Both constant groups trained with three repetitions of the same word scramble, but for constant group A, the testing phase consisted of the identical letter arrangement to that seen during training (e.g. LDOOF), whereas for constant group B, the testing phase consisted of a arrangement they had not seen during training, thus presenting them with a testing situation similar situation to the varied group. At the testing stage, the varied group outperformed both constant groups, a particularly impressive result, given that constant group A had three prior exposures to the word arrangement (i.e. the particular permutation of letters) which the varied group had not explicitly seen. However varied subjects in this study did not exhibit the typical decrement in the training phase typical of other varied manipulations in the literature, and actually achieved higher levels of anagram solving accuracy by the end of training than either of the constant groups -- solving two more anagrams on average than the constant group. This might suggest that for tasks of this nature where the learner can simply get stuck with a particular word scramble, repeated exposure to the identical scramble might be less helpful towards finding the solution than being given a different arrangement of the same letters. This contention is supported by the fact that constant group A, who was tested on the identical arrangement as they experienced during training, performed no better at testing than did constant group B, who had trained on a different arrangement of the same word solution -- further suggesting that there may not have been a strong identity advantage in this task.

@@

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g. the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 conducted two experiments with this basic paradigm. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into the category with greater variation.

@@

In an example of a more complex influence of training variation, [@braithwaiteEffectsVariationPrior2015] trained participants on example problems involving the concept of sampling with replacement (SWR). Training consisted of examples that were either highly similar in their semantic context (e.g. all involving people selecting objects) or in which the surface features were varied between examples (e.g. people choosing objects AND objects selected in a sequence). The experimenters also surveyed how much prior knowledge each participant had with SWR. They found that whether variation was beneficial depended on the prior knowledge of the participants -- such that participants with some prior knowledge benefited from varied training, whereas participants with minimal prior knowledge performed better after training with similar examples. The authors hypothesized that in order to benefit from varied examples, participants must be able to detect the structure common to the diverse examples, and that participants with prior knowledge are more likely to be sensitive to such structure, and thus to benefit from varied training. To test this hypothesis more directly, the authors conducted a 2nd experiment, wherein they controlled prior knowledge by exposing some subjects to a short graphical or verbal pre-training lesson, designed to increase sensitivity to the training examples. Consistent with their hypothesis, participants exposed to the structural sensitivity pre-training benefited more from varied training than the controls participants who benefited more from training with similar examples. Interactions between prior experience and the influence of varied training have also been observed in sensorimotor learning [@guadagnoliRelationshipContextualInterference1999;  @delreyEffectsContextualInterference1982].  @delreyEffectsContextualInterference1982 recruited participants who self-reported either extensive, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, with one of several varied training procedures. Unsurprisingly, athlete participants had superior performance during training, regardless of  condition, and training performance was superior for all subjects in the constant group. Of greater interest is the pattern of testing results from novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received variable training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training.

@@

A number of theoretical frameworks have been proposed to conceptually explain the effects of varied training on learning and generalization. Schema theory (described in more detail above), posts that varied practice leads to the formation of more flexible motor schemas, which then facilitate generalization [@schmidtSchemaTheoryDiscrete1975]. The desirable difficulties framework [@bjorkMakingThingsHard2011; @soderstromLearningPerformanceIntegrative2015] proposes that variable practice conditions may impair initial performance but then enhance longer-term retention and transfer. Similarily, the challenge point framework [@guadagnoliChallengePointFramework2004] contends that training variation induces optimal learning occurs insofar as it causes the difficulty of practice tasks to be appropriately matched to the learner's capabilities, but may also be detrimental if the amount of variation causes the task to be too difficult.

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited as evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of the "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992; @knappTheoryCategorizationBased1984]. @palmeriCentralTendenciesExtreme2001 demonstrated both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

@@

Although the benefits of training variation in visuomotor skill learning have been observed many times, null findings have also been repeatedly found, leading some researchers to question the veracity of the variability of practice hypothesis [@newellSchemaTheory19752003; @vanrossumSchmidtSchemaTheory1990]. Critics have also pointed out that investigations of the effects of training variability, of the sort described above, often fail to control for the effect of similarity between training and testing conditions. For training tasks in which participants have numerous degrees of freedom (e.g. projectile throwing tasks where participants control the x and y velocity of the projectile), varied groups are likely to experience a wider range of the task space over the course of their training (e.g. more unique combinations of x and y velocities).
Experimenters may attempt to account for this possibility by ensuring that the training location(s) of the varied and constant groups are an equal distance away from the eventual transfer locations, such that their training throws are, on average, equally similar to throws that would lead to good performance at the transfer locations. However, even this level of experimental control may still be insufficient to rule out the effect of similarity on transfer. Given that psychological similarity is typically best described as either a Gaussian or exponentially decaying function of psychological distance [@ennisMultidimensionalStochasticTheory1988; @ghahramaniGeneralizationLocalRemappings1996; @loganInstanceTheoryAutomatization1988; @nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987; @thoroughmanRapidReshapingHuman2005], it is plausible that a subset of the most similar training instances could have a disproportionate impact on generalization to transfer conditions, even if the average distance between training and transfer conditions is identical between groups. @fig-toy-model1 demonstrates the consequences of a generalization gradient that drops off as a Gaussian function of distance from training, as compared to a linear drop-off.

@@

Participants were recruited from an undergraduate population that is 63% female and consists almost entirely of individuals aged 18 to 22 years. A total of 110 Indiana University psychology students participated in Experiment 1. We subsequently excluded 34 participants for poor performance on one of the dependent measures of the task (2.5-3 standard deviations worse than the median subject at the task) or for displaying a pattern of responses that was clearly indicative of a lack of engagement with the task (e.g. simply dropping the ball on each trial rather than throwing it at the target), or for reporting that they completed the experiment on a phone or tablet device, despite the instructions not to use one of these devices. A total of 74 participants were retained for the final analyses, 35 in the varied group and 39 in the constant group.

@@

```{r}
#| label: tbl-e2table2
#| tbl-cap: Testing performance from novel positions. Includes data only from positions that were not encountered during the training stage (e.g. excludes positions 500 and 800 for the varied group, and one of the six locations for each of the constant groups). Table presents Mean absolute deviations from the center of the target, and standard deviations in parenthesis.

@@

```{r}
#| label: fig-e2test1
#| fig.cap: A comparison of throwing location that are identical to those trained by the constant participants (e.g. constant participants trained at position 900, tested from position 900), which are also novel to the varied-trained participants (thus excluding positions 500 and 800). Error bars indicate standard error of the mean.

@@

It is common for psychological process models of categorization learning to use an approach such as multidimensional scaling so as to transform the stimuli from the physical dimensions used in the particular task into the psychological dimensions more reflective of the actual human representations [@nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987]. Such scaling typically entails having participants rate the similarity between individual items and using these similarity judgements to then compute the psychological distances between stimuli, which can then be fed into a subsequent model. In the present investigation, there was no such way to scale the x and y velocity components in terms of the psychological similarity, and thus our modelling does rely on the assumption that the psychological distances between the different throwing positions are proportional to absolute distances in the metric space of the task (e.g. the relative distance between positions 400 and 500 is equivalent to that between 800 and 900). However, an advantage of our approach is that we are measuring similarity in terms of how participants behave (applying a velocity to the ball), rather than the metric features of the task stimuli.

>>>>

Varied training has been shown to influence learning in a wide array of different tasks and domains, including categorization [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968; @nosofskyModelguidedSearchOptimal2019; @morgensternOneshotCategorizationNovel2019; @plebanekEffectsFrequencyVariability2021], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012; @brekelmansDoesHighVariability2022], anagram completion [@goodeSuperiorityVariableRepeated2008], perceptual learning [@manentiVariabilityTrainingUnlocks2023; @lovibondStimulusDiscriminabilityInduction2020; @robsonSpecificVariedPractice2022a; @zamanPerceptualVariabilityImplications2021], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], cognitive control tasks [@sabahWhenLessMore2019; @moshon-cohenStimulusVariabilityImproves2024], associative learning [@leeEvidentialDiversityIncreases2019; @reichmannVariabilityAbstractionEvaluative2023; @fanStimulusDiversityIncreases2022; @pradaExperiencedCategoryVariability2020; @liveseyRevisitingPeakShift2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], face recognition [@honigPerceptualSimilarityModulates2022; @burtonIdentityVariationRepresentations2016; @menonVariationPhotosSame2015],  the perception of social group heterogeneity [@linvilleExemplarAbstractionModels1993; @konovalovaInformationSamplingExplanation2020; @parkPerceptionVariabilityCategory1987; @gershmanStructureLearningPrinciples2023] ,  simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019; @breslinConstantVariablePractice2012], and  complex skill learning [@seowTransferEffectsVaried2019; @huetEducationAttentionExplanation2011; @hacquesVisualControlClimbing2022]. See @czyzVariabilityPracticeInformation2021 or @ravivHowVariabilityShapes2022 for more detailed reviews.

Research on the effects of varied training typically manipulates variability in one of two ways. In the first approach, a high variability group is exposed to a greater number of unique instances during training, while a low variability group receives fewer unique instances with more repetitions. Alternatively, both groups may receive the same number of unique instances, but the high variability group's instances are more widely distributed or spread out in the relevant psychological space, while the low variability group's instances are clustered more tightly together. Researchers then compare the training groups in terms of their performance during the training phase, as well as their generalization performance during a testing phase. Researchers usually compare the performance of the two groups during both the training phase and a subsequent testing phase. The primary theoretical interest is often to assess the influence of training variability on generalization to novel testing items or conditions. However, the test may also include some or all of the items that were used during the training stage, allowing for an assessment of whether the variability manipulation influenced the learning of the trained items themselves, or to easily measure how much performance degrades as a function of how far away testing items are from the training items.

The influence of training variability has received a large amount of attention in the domain of sensorimotor skill learning. Much of this research has been influenced by the work of @schmidtSchemaTheoryDiscrete1975, who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. Schema theory presumes that learners possess general motor programs for a class of movements (e.g., an underhand throw). When called up for use motor programs are parameterized by schema rules which determine how the motor program is parameterized or scaled to the particular demands of the current task. Schema theory predicts that variable training facilitates the formation of more robust schemas, which will result in improved generalization or transfer. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices with multiple instances along some task-relevant dimension that remains invariant for the constant group. For example, studies using a projectile throwing task might assign participants to either constant training that practice throwing from a single location, or to a varied group that throws from multiple locations. Following training, both groups are then tested from novel throwing locations [@pigottMotorSchemaStructure1984; @willeyLimitedGeneralizationVaried2018; @pachecoLearningSpecificIndividual2018; @wulfEffectTypePractice1991].

One of the earliest and still often cited investigations of Schmidt's benefits of variability hypothesis was the work of @kerrSpecificVariedPractice1978. Two groups of children, aged 8 and 12, were assigned to either constant or varied training of a bean bag throwing task. The constant group practiced throwing a bean-bag at a small target placed 3 feet in front of them, and the varied group practiced throwing from a distance of both 2 feet and 4 feet. Participants were blindfolded and unable to see the target while making each throw but would receive feedback by looking at where the beanbag had landed in between each training trial. 12 weeks later, all of the children were given a final test from a distance of 3 feet which was novel for the varied participants and repeated for the constant participants. Participants were also blindfolded for testing and did not receive trial by trial feedback in this stage. In both age groups, participants performed significantly better in the varied condition than the constant condition, though the effect was larger for the younger, 8-year-old children. This result provides particularly strong evidence for the benefits of varied practice, as the varied group outperformed the constant group even when tested at the "home-turf" distance that the constant group had exclusively practiced. A similar pattern of results was observed in another study wherein varied participants trained with tennis, squash, badminton, and short-tennis rackets were compared against constant subjects trained with only a tennis racket [@greenPracticeVariabilityTransfer1995a]. One of the testing conditions had subjects repeat the use of the tennis racket, which had been used on all 128 training trials for the constant group, and only 32 training trials for the varied group. Nevertheless, the varied group outperformed the constant group when using the tennis racket at testing, and also performed better in conditions with several novel racket lengths. However, as is the case with many of the patterns commonly observed in the "benefits of variability" literature, the pattern wherein the varied group outperfroms the constant group even from the constants group's home turf has not been consistently replicated. One recent study attempted a near replication of the  Kerr & Booth study [@willeyLongtermMotorLearning2018], having subjects throw beanbags at a target, with the varied group training from positions (5 and 9 feet) on either side of the constant group (7 feet). This study did not find a varied advantage from the constant training position, though the varied group did perform better at distances novel to both groups. However, this study diverged from the original in that the participants were adults; and the amount of training was much greater (20 sessions with 60 practice trials each, spread out over 5-7 weeks).

@@

Benefits of varied training have also been observed in many studies outside of the sensorimotor domain.  @goodeSuperiorityVariableRepeated2008 trained participants to solve anagrams of 40 different words ranging in length from 5 to 11 letters, with an anagram of each word repeated 3 times throughout training, for a total of 120 training trials. Although subjects in all conditions were exposed to the same 40 unique words (i.e. the solution to an anagram), participants in the varied group saw 3 different arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for the solution word FLOOD, whereas constant subjects would train on three repetitions of LDOOF (spread evenly across training). Two different constant groups were used. Both constant groups trained with three repetitions of the same word scramble, but for constant group A, the testing phase consisted of the identical letter arrangement to that seen during training (e.g., LDOOF), whereas for constant group B, the testing phase consisted of a arrangement they had not seen during training, thus presenting them with a testing situation similar situation to the varied group. At the testing stage, the varied group outperformed both constant groups, a particularly impressive result, given that constant group A had three prior exposures to the word arrangement (i.e. the particular permutation of letters) which the varied group had not explicitly seen. However varied subjects in this study did not exhibit the typical decrement in the training phase typical of other varied manipulations in the literature, and actually achieved higher levels of anagram solving accuracy by the end of training than either of the constant groups -- solving two more anagrams on average than the constant group. This might suggest that for tasks of this nature where the learner can simply get stuck with a particular word scramble, repeated exposure to the identical scramble might be less helpful towards finding the solution than being given a different arrangement of the same letters. This contention is supported by the fact that constant group A, who was tested on the identical arrangement as they experienced during training, performed no better at testing than did constant group B, who had trained on a different arrangement of the same word solution -- further suggesting that there may not have been a strong identity advantage in this task.

@@

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g., the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 conducted two experiments with this basic paradigm. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into the category with greater variation.

@@

In an example of a more complex influence of training variation, [@braithwaiteEffectsVariationPrior2015] trained participants on example problems involving the concept of sampling with replacement (SWR). Training consisted of examples that were either highly similar in their semantic context (e.g., all involving people selecting objects) or in which the surface features were varied between examples (e.g., people choosing objects AND objects selected in a sequence). The experimenters also surveyed how much prior knowledge each participant had with SWR. They found that whether variation was beneficial depended on the prior knowledge of the participants -- such that participants with some prior knowledge benefited from varied training, whereas participants with minimal prior knowledge performed better after training with similar examples. The authors hypothesized that in order to benefit from varied examples, participants must be able to detect the structure common to the diverse examples, and that participants with prior knowledge are more likely to be sensitive to such structure, and thus to benefit from varied training. To test this hypothesis more directly, the authors conducted a 2nd experiment, wherein they controlled prior knowledge by exposing some subjects to a short graphical or verbal pre-training lesson, designed to increase sensitivity to the training examples. Consistent with their hypothesis, participants exposed to the structural sensitivity pre-training benefited more from varied training than the controls participants who benefited more from training with similar examples. Interactions between prior experience and the influence of varied training have also been observed in sensorimotor learning [@guadagnoliRelationshipContextualInterference1999;  @delreyEffectsContextualInterference1982].  @delreyEffectsContextualInterference1982 recruited participants who self-reported either extensive, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, with one of several varied training procedures. Unsurprisingly, athlete participants had superior performance during training, regardless of  condition, and training performance was superior for all subjects in the constant group. Of greater interest is the pattern of testing results from novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received variable training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training.

@@

A number of theoretical frameworks have been proposed to conceptually explain the effects of varied training on learning and generalization. Schema theory (described in more detail above), posts that varied practice leads to the formation of more flexible motor schemas, which then facilitate generalization [@schmidtSchemaTheoryDiscrete1975]. The desirable difficulties framework [@bjorkMakingThingsHard2011; @soderstromLearningPerformanceIntegrative2015] proposes that variable practice conditions may impair initial performance but then enhance longer-term retention and transfer. Similarly, the challenge point framework [@guadagnoliChallengePointFramework2004] contends that training variation induces optimal learning occurs insofar as it causes the difficulty of practice tasks to be appropriately matched to the learner's capabilities, but may also be detrimental if the amount of variation causes the task to be too difficult.

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.,  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited as evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of the "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992; @knappTheoryCategorizationBased1984]. @palmeriCentralTendenciesExtreme2001 demonstrated both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

@@

Although the benefits of training variation in visuomotor skill learning have been observed many times, null findings have also been repeatedly found, leading some researchers to question the veracity of the variability of practice hypothesis [@newellSchemaTheory19752003; @vanrossumSchmidtSchemaTheory1990]. Critics have also pointed out that investigations of the effects of training variability, of the sort described above, often fail to control for the effect of similarity between training and testing conditions. For training tasks in which participants have numerous degrees of freedom (e.g., projectile throwing tasks where participants control the x and y velocity of the projectile), varied groups are likely to experience a wider range of the task space over the course of their training (e.g., more unique combinations of x and y velocities).
Experimenters may attempt to account for this possibility by ensuring that the training location(s) of the varied and constant groups are an equal distance away from the eventual transfer locations, such that their training throws are, on average, equally similar to throws that would lead to good performance at the transfer locations. However, even this level of experimental control may still be insufficient to rule out the effect of similarity on transfer. Given that psychological similarity is typically best described as either a Gaussian or exponentially decaying function of psychological distance [@ennisMultidimensionalStochasticTheory1988; @ghahramaniGeneralizationLocalRemappings1996; @loganInstanceTheoryAutomatization1988; @nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987; @thoroughmanRapidReshapingHuman2005], it is plausible that a subset of the most similar training instances could have a disproportionate impact on generalization to transfer conditions, even if the average distance between training and transfer conditions is identical between groups. @fig-toy-model1 demonstrates the consequences of a generalization gradient that drops off as a Gaussian function of distance from training, as compared to a linear drop-off.

@@

Participants were recruited from an undergraduate population that is 63% female and consists almost entirely of individuals aged 18 to 22 years. A total of 110 Indiana University psychology students participated in Experiment 1. We subsequently excluded 34 participants for poor performance on one of the dependent measures of the task (2.5-3 standard deviations worse than the median subject at the task) or for displaying a pattern of responses that was clearly indicative of a lack of engagement with the task (e.g., simply dropping the ball on each trial rather than throwing it at the target), or for reporting that they completed the experiment on a phone or tablet device, despite the instructions not to use one of these devices. A total of 74 participants were retained for the final analyses, 35 in the varied group and 39 in the constant group.

@@

```{r}
#| label: tbl-e2table2
#| tbl-cap: Testing performance from novel positions. Includes data only from positions that were not encountered during the training stage (e.g., excludes positions 500 and 800 for the varied group, and one of the six locations for each of the constant groups). Table presents Mean absolute deviations from the center of the target, and standard deviations in parenthesis.

@@

```{r}
#| label: fig-e2test1
#| fig.cap: A comparison of throwing location that are identical to those trained by the constant participants (e.g., constant participants trained at position 900, tested from position 900), which are also novel to the varied-trained participants (thus excluding positions 500 and 800). Error bars indicate standard error of the mean.

@@

It is common for psychological process models of categorization learning to use an approach such as multidimensional scaling so as to transform the stimuli from the physical dimensions used in the particular task into the psychological dimensions more reflective of the actual human representations [@nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987]. Such scaling typically entails having participants rate the similarity between individual items and using these similarity judgements to then compute the psychological distances between stimuli, which can then be fed into a subsequent model. In the present investigation, there was no such way to scale the x and y velocity components in terms of the psychological similarity, and thus our modelling does rely on the assumption that the psychological distances between the different throwing positions are proportional to absolute distances in the metric space of the task (e.g., the relative distance between positions 400 and 500 is equivalent to that between 800 and 900). However, an advantage of our approach is that we are measuring similarity in terms of how participants behave (applying a velocity to the ball), rather than the metric features of the task stimuli.
